---
title: "Package Management with renv"
author: "Akash B V"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 2
---

## Overview

This project uses [renv](https://rstudio.github.io/renv/) to manage R package dependencies. The lockfile (`renv.lock`) records exact package versions, ensuring reproducibility across machines and collaborators.

**Key files:**

| File | Purpose |
|------|---------|
| `renv.lock` | Records exact package versions (committed to git) |
| `renv/activate.R` | Bootstraps renv on session start |
| `.Rprofile` | Calls `activate.R` automatically |

## Setup

### First-time setup

```bash
# Clone repository
git clone https://github.com/ccmmf/uncertainty.git
cd uncertainty

# Create your .Renviron (required before renv works on SCC)
cp .Renviron.example .Renviron
# Edit paths if needed

# Restore packages from lockfile
R -e "renv::restore()"
```

### Verifying installation

```{r}
#| eval: false
# Check that PEcAn packages load
library(PEcAn.settings)
library(PEcAn.uncertainty)
library(sensobol)
```

## Common operations

### Adding packages

```{r}
#| eval: false
# Install from CRAN
install.packages("newpackage")

# Install from GitHub (e.g. PEcAn development branch)
renv::install("PecanProject/pecan@develop", subdir = "modules/uncertainty")

# Update lockfile
renv::snapshot()
```

### After pulling changes

If `renv.lock` changed in a pull:

```{r}
#| eval: false
renv::restore()
```

### Checking status

```{r}
#| eval: false
# Compare library state to lockfile
renv::status()

# See which packages are used in project code
renv::dependencies()
```

## SCC-specific notes

### Library location

On BU SCC, home directories have limited quota. The `.Renviron` file redirects renv's cache and library to the shared project space:
```
RENV_PATHS_CACHE=/projectnb/dietzelab/ccmmf/renv-cache
RENV_PATHS_LIBRARY=/projectnb/dietzelab/ccmmf/renv-library
```

This must be set **before** running `renv::restore()`.

### Module dependencies

Before starting R, load required system modules:

```bash
module load R/4.4.0 gdal proj geos sqlite udunits
```

## Troubleshooting

**"there is no package called 'renv'"**

Install renv globally first:
```{r}
#| eval: false
install.packages("renv")
```

**"cannot allocate vector of size X Gb"**

renv cache is in home directory (quota exceeded). Set `RENV_PATHS_CACHE` in `.Renviron`.

**Package not found in repositories**

PEcAn packages require the R-universe repo. Check that `.Rprofile` includes:
```{r}
#| eval: false
options(repos = c(
  PEcAn = "https://pecanproject.r-universe.dev",
  CRAN  = "https://cloud.r-project.org"
))
```

**Wrong package version loading**

Check library search path:
```{r}
#| eval: false
.libPaths()  # Project library should be first
```

If system library is first, ensure `renv/activate.R` is being sourced in `.Rprofile`.

## References

- [renv documentation](https://rstudio.github.io/renv/)
- [PEcAn R-universe](https://pecanproject.r-universe.dev/)