---
title: "Global Sensitivity Analysis Using Sobol Indices"
author: "Akash B V"
date: last-modified
format:
  html:
    self-contained: true
    df-print: paged
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    theme: cosmo
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

# Overview

This report presents results from a **variance-based global sensitivity analysis** of the SIPNET ecosystem model using Sobol indices. Unlike local sensitivity analysis (which measures rates of change at baseline conditions), variance-based methods decompose output variance into contributions from individual parameters and their interactions.

::: callout-note
## Why Sobol Indices?

Sobol indices provide a rigorous framework for:

-   **Quantifying parameter importance** via variance decomposition
-   **Detecting interactions** between parameters (non-additive effects)
-   **Factor prioritization** (which parameters to constrain for uncertainty reduction)
-   **Factor fixing** (which parameters can be fixed without loss of information)
:::

## Theoretical Background

**Sobol indices** are normalized variance contributions:

$$
S_i = \frac{V_i}{\text{VAR}(y)} \quad \text{(First-order index)}
$$ {#eq-si}

$$
T_i = 1 - \frac{\text{VAR}_{x_{\sim i}}[\mathbb{E}_{x_i}(y | x_{\sim i})]}{\text{VAR}(y)} = \frac{\mathbb{E}_{x_{\sim i}}[\text{VAR}_{x_i}(y | x_{\sim i})]}{\text{VAR}(y)} \quad \text{(Total-order index)}
$$ {#eq-ti}

where $x_{\sim i}$ denotes all-parameters-but-$x_i$.

### Interpretation

-   $S_i$: Direct effect of $x_i$ on output variance (excluding interactions)
-   $T_i$: Total effect of $x_i$ including all interactions with other parameters
-   $T_i - S_i$: Interaction strength (purely higher-order effects)
-   $\sum S_i \approx 1$: Model is **additive** (linear-like behavior)
-   $\sum S_i \ll 1$: Model is **non-additive** (interactions dominate)

# Study Design

```{r}
#| label: setup
#| include: false

library(here)
here::i_am("analysis/global_sensitivity.qmd")

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(knitr)
library(scales)
library(config)

cfg <- config::get(file = here::here("000-config.yml"))

sobol_indices <- readr::read_csv(
  here::here(cfg$paths$data_dir, "sobol_indices.csv"),
  show_col_types = FALSE
)
metadata <- readRDS(here::here(cfg$paths$data_dir, "sobol_design_matrix_metadata.rds"))

N <- metadata$N
k <- metadata$k
total_runs <- metadata$total_runs
n_sites <- dplyr::n_distinct(sobol_indices$runid)
response_vars <- unique(sobol_indices$variable)

# consistent theme
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      strip.text = element_text(size = 10, margin = margin(2, 2, 2, 2)),
      legend.position = "bottom"
    )
)
```

**Method:** Saltelli's sampling scheme with Jansen estimators

**Sample Size:** $N = `r N`$, Total Runs = $N \times (k + 2)$ = `r total_runs`

**Scope:** `r n_sites` sites across California row-crop agriculture

# Data Processing

We process the raw indices to parse namespaced parameters (e.g. `grass.SLA`), identify the Dummy parameter (Sobol validation -- its $T_i$ should be near zero), and calculate interaction components ($T_i - S_i$).

```{r}
#| label: data-processing

# reshape and enforce non-negativity (numerical artifacts can cause small negatives)
sobol_long <- sobol_indices |>
  select(runid, variable, parameters,
         Si_original, Si_low.ci, Si_high.ci,
         Ti_original, Ti_low.ci, Ti_high.ci) |>
  mutate(
    Si = pmax(Si_original, 0),
    Ti = pmax(Ti_original, 0),
    Interaction = pmax(Ti - Si, 0)
  )

# parse parameter names: "pft.param" -> Prefix + Trait
sobol_parsed <- sobol_long |>
  separate(parameters, into = c("Prefix", "Trait"), sep = "\\.",
           extra = "merge", fill = "right") |>
  mutate(
    Trait = ifelse(is.na(Trait), Prefix, Trait),
    Category = case_when(
      Prefix == "ic_ensemble" ~ "Initial Conditions",
      Prefix == "met_ensemble" ~ "Meteorology",
      Prefix == "dummy" ~ "Control (Noise)",
      Prefix == "mgmt" ~ "Management",
      TRUE ~ "Biological Parameter"
    ),
    Label = case_when(
      Prefix == "ic_ensemble" ~ "Init. Cond.",
      Prefix == "met_ensemble" ~ "Met. Driver",
      Prefix == "dummy" ~ "Dummy (Noise)",
      Prefix == "mgmt" ~ str_to_title(gsub("_", " ", Trait)),
      TRUE ~ paste(str_to_title(Prefix), Trait)
    )
  )

# per-site summaries (keep distribution for cross-site variability)
sobol_site <- sobol_parsed |>
  summarize(
    Si = mean(Si, na.rm = TRUE),
    Ti = mean(Ti, na.rm = TRUE),
    Interaction = mean(Interaction, na.rm = TRUE),
    .by = c("runid", "variable", "Category", "Label")
  )

# cross-site summary: median + IQR
sobol_summary <- sobol_site |>
  summarize(
    median_Ti = median(Ti, na.rm = TRUE),
    q25_Ti = quantile(Ti, 0.25, na.rm = TRUE),
    q75_Ti = quantile(Ti, 0.75, na.rm = TRUE),
    median_Si = median(Si, na.rm = TRUE),
    median_Int = median(Interaction, na.rm = TRUE),
    .by = c("variable", "Category", "Label")
  )
```

# Key Findings

```{r}
#| label: key-findings
#| results: asis

# additivity
additivity <- sobol_parsed |>
  filter(Category != "Control (Noise)") |>
  summarize(Sum_Si = sum(Si, na.rm = TRUE), .by = c("runid", "variable")) |>
  summarize(mean_sum = mean(Sum_Si), .by = "variable")

# top 3 parameters
top3 <- sobol_summary |>
  filter(Category != "Control (Noise)") |>
  arrange(desc(median_Ti)) |>
  slice_head(n = 3)

# biology fraction
bio_frac <- sobol_parsed |>
  filter(Category != "Control (Noise)") |>
  summarize(Total_Si = sum(Si, na.rm = TRUE), .by = c("variable", "Category")) |>
  mutate(frac = Total_Si / sum(Total_Si), .by = "variable") |>
  filter(Category == "Biological Parameter") |>
  pull(frac) |>
  mean()

# noise floor
dummy_vals <- sobol_summary |>
  filter(Category == "Control (Noise)") |>
  pull(median_Ti) |>
  mean()
```

Based on a Sobol analysis with `r N` base samples and `r total_runs` total runs across `r n_sites` sites:

-   **Top parameters**: `r top3$Label[1]`, `r top3$Label[2]`, and `r top3$Label[3]` explain the most output variance ($T_i$ = `r round(top3$median_Ti[1], 2)`, `r round(top3$median_Ti[2], 2)`, `r round(top3$median_Ti[3], 2)`)
-   **Biological parameters** account for \~`r round(bio_frac * 100)`% of explained variance; meteorological and initial condition drivers explain the rest
-   **Model additivity** ($\sum S_i$) averages `r round(mean(additivity$mean_sum), 2)` -- values near 1 indicate additive behavior, values $\ll 1$ indicate parameter interactions dominate
-   **Noise floor**: the dummy parameter's median $T_i$ = `r round(dummy_vals, 4)`, providing a rigorous threshold for identifying non-influential inputs

# Variance Decomposition (Source Partitioning) {#sec-variance-partition}

Before ranking individual parameters, we assess the *sources* of uncertainty. We sum the First-Order indices ($S_i$) by category (see @eq-si). The "Control (Noise)" segment is excluded to avoid confusion; it represents numerical estimation error from the Sobol design.

```{r}
#| label: fig-variance-partition
#| fig-cap: "Proportion of variance explained by Biology (parameters) vs. Environment (IC / met). This partitioning answers: is output uncertainty dominated by what we know about the plants, or by the weather and initial conditions?"
#| fig-width: 8
#| fig-height: 5

variance_partition <- sobol_parsed |>
  filter(Category != "Control (Noise)") |>
  summarize(Total_Si = sum(Si, na.rm = TRUE), .by = c("variable", "Category")) |>
  mutate(Percentage = Total_Si / sum(Total_Si), .by = "variable")

ggplot(variance_partition, aes(x = variable, y = Total_Si, fill = Category)) +
  geom_col(position = "fill", width = 0.7) +
  geom_text(aes(label = percent(Percentage, accuracy = 1)),
            position = position_fill(vjust = 0.5),
            color = "white", fontface = "bold", size = 3.5) +
  scale_y_continuous(labels = percent) +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    y = "Proportion of Variance",
    x = NULL,
    fill = "Source"
  ) +
  coord_flip()
```

# Parameter Ranking (First & Total Order) {#sec-rankings}

This plot compares First-Order ($S_i$, solid bar) and Total-Order ($T_i$, whisker) indices. The gap between them quantifies interaction strength (@eq-ti minus @eq-si).

**Solid bar** ($S_i$): Direct (main) effect of the parameter alone

**Gap** ($T_i - S_i$): Higher-order interaction effects with other parameters

Parameters are ordered by median $T_i$ across sites. We show the top 15 per output variable.

```{r}
#| label: fig-ranking-combined
#| fig-cap: "Top 15 inputs ranked by median Total Effect (Ti) across sites. Solid bar = Main Effect (Si); whisker = Total Effect (Ti). Parameters with large gaps between Si and Ti interact strongly with other inputs."
#| fig-height: 10
#| fig-width: 10

top_params <- sobol_summary |>
  filter(Category != "Control (Noise)") |>
  slice_max(order_by = median_Ti, n = 15, by = "variable")

ggplot(top_params, aes(x = reorder(Label, median_Ti), y = median_Ti)) +
  geom_errorbar(
    aes(ymin = median_Si, ymax = median_Ti),
    width = 0.2, color = "grey50"
  ) +
  geom_point(aes(y = median_Ti, color = Category), shape = 95, size = 10) +
  geom_col(aes(y = median_Si, fill = Category), alpha = 0.8) +
  coord_flip() +
  facet_wrap(~variable, scales = "free_x") +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Set2") +
  labs(
    x = NULL,
    y = "Sobol Index (median across sites)"
  )
```

## Cross-Site Variability

A parameter whose importance varies across sites (wide box) indicates context-dependent sensitivity -- its influence depends on local conditions (soil, climate, crop type). These are priorities for spatially explicit calibration.

```{r}
#| label: fig-site-variability
#| fig-cap: "Distribution of Ti across sites for top 10 parameters. Points = site-level values; box = IQR; line = median. Wide boxes indicate sensitivity varies with local conditions."
#| fig-height: 8
#| fig-width: 10

# top 10 by median Ti (across all variables)
top10_labels <- sobol_summary |>
  filter(Category != "Control (Noise)") |>
  slice_max(order_by = median_Ti, n = 10) |>
  pull(Label) |>
  unique()

sobol_site |>
  filter(Label %in% top10_labels, Category != "Control (Noise)") |>
  ggplot(aes(x = reorder(Label, Ti, FUN = median), y = Ti)) +
  geom_boxplot(aes(fill = Category), outlier.size = 0.8, alpha = 0.7) +
  coord_flip() +
  facet_wrap(~variable, scales = "free_x") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = NULL,
    y = "Total-order index (Ti)",
    fill = "Source"
  )
```

# Interaction Analysis {#sec-interactions}

Parameters with large differences between $T_i$ and $S_i$ interact strongly with other factors. These non-additive effects mean the parameter's impact on output depends on the values of other inputs.

```{r}
#| label: fig-interactions
#| fig-cap: "Interaction strength (Ti - Si) for parameters with interactions exceeding 5% of total variance. Numbers show percent of Ti attributable to interactions."
#| fig-height: 6
#| fig-width: 9

high_int <- sobol_summary |>
  filter(Category != "Control (Noise)", median_Int > 0.05) |>
  mutate(Int_pct = (median_Int / median_Ti) * 100)

if (nrow(high_int) > 0) {
  ggplot(high_int, aes(x = reorder(Label, median_Int), y = median_Int)) +
    geom_col(aes(fill = Category), alpha = 0.8) +
    geom_text(
      aes(label = paste0(round(Int_pct, 0), "%")),
      hjust = -0.1, size = 3
    ) +
    coord_flip() +
    facet_wrap(~variable, scales = "free_x") +
    scale_fill_brewer(palette = "Set2") +
    labs(
      x = NULL,
      y = "Interaction strength (Ti - Si)",
      fill = "Source"
    )
} else {
  cat("No parameters exceeded the 5% interaction threshold.")
}
```

# Factor Fixing (Screening) {#sec-screening}

To identify non-influential parameters, we compare their Total-Order index ($T_i$) against the **Dummy parameter** (noise floor). The dummy parameter varies randomly but has no effect on model output, so any parameter with $T_i$ indistinguishable from the dummy can be safely fixed to a constant value without meaningful loss of variance coverage.

The dashed red line marks the noise floor (median dummy $T_i$).

```{r}
#| label: fig-screening
#| fig-cap: "Factor screening: parameters below the noise line (dashed red) can be fixed. Dot = median Ti across sites; whisker = IQR."
#| fig-height: 16
#| fig-width: 12

dummy_thresholds <- sobol_summary |>
  filter(Category == "Control (Noise)") |>
  select(variable, noise_Ti = median_Ti)

screening <- sobol_summary |>
  filter(Category != "Control (Noise)") |>
  left_join(dummy_thresholds, by = "variable")

ggplot(screening, aes(x = reorder(Label, median_Ti), y = median_Ti)) +
  geom_point(aes(color = Category), size = 2.5) +
  geom_errorbar(
    aes(ymin = q25_Ti, ymax = q75_Ti, color = Category),
    width = 0.3, alpha = 0.6
  ) +
  geom_hline(
    data = dummy_thresholds,
    aes(yintercept = noise_Ti),
    linetype = "dashed", color = "red", alpha = 0.7
  ) +
  coord_flip() +
  facet_wrap(~variable, scales = "free_x") +
  scale_color_brewer(palette = "Set2") +
  labs(
    x = NULL,
    y = "Total-order index (Ti)",
    color = "Source"
  ) +
  theme(axis.text.y = element_text(size = 8))
```

```{r}
#| label: fixable-params
#| results: asis

fixable <- screening |>
  filter(median_Ti <= noise_Ti) |>
  select(Variable = variable, Input = Label,
         `Median Ti` = median_Ti, `Noise Threshold` = noise_Ti) |>
  arrange(Variable, `Median Ti`)

if (nrow(fixable) > 0) {
  cat("\n**Fixable parameters** (Ti below noise floor):\n\n")
  knitr::kable(fixable, digits = 4)
} else {
  cat("\n**No parameters met the screening criteria.** All inputs exceed the numerical noise floor.\n")
}
```

# Model Additivity {#sec-additivity}

The sum of First-Order indices ($\sum S_i$) indicates the degree of model linearity. Values near 1.0 imply the model is **additive** (parameters contribute independently). Values well below 1 indicate that **interactions dominate** -- the model's response to one parameter depends on the values of others.

```{r}
#| label: fig-additivity
#| fig-cap: "Distribution of model additivity (sum of Si) across sites. Dashed line at 1.0 marks perfect additivity. Values below 1 indicate interaction effects."
#| fig-height: 4
#| fig-width: 7

additivity_site <- sobol_parsed |>
  filter(Category != "Control (Noise)") |>
  summarize(Sum_Si = sum(Si, na.rm = TRUE), .by = c("runid", "variable"))

ggplot(additivity_site, aes(x = variable, y = Sum_Si)) +
  geom_boxplot(fill = "steelblue", alpha = 0.5, outlier.size = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  labs(x = NULL, y = expression(sum(S[i]))) +
  coord_flip()
```

# Synthesis

**What drives uncertainty?** The variance decomposition (@sec-variance-partition) separates biological parameter uncertainty from environmental forcing.

**What to calibrate first?** Parameter rankings (@sec-rankings) identify the top candidates for reducing predictive uncertainty. Parameters whose importance varies across sites (@fig-site-variability) need spatially explicit calibration.

**What can be simplified?** Factor fixing (@sec-screening) uses the dummy parameter as a rigorous noise floor to identify inputs that can be set to nominal values.

**Is the model additive?** Additivity diagnostics (@sec-additivity) indicate whether parameter effects are independent or interact. Strong interactions (large $T_i - S_i$ in @sec-interactions) suggest that calibrating parameters independently may be insufficient.

# References

1.  Puy, A., et al. (2022). sensobol: An R Package to Compute Variance-Based Sensitivity Indices. *Journal of Statistical Software*, 102(5).

2.  Saltelli, A., et al. (2010). Variance based sensitivity analysis of model output. *Computer Physics Communications*.

3.  Dietze, M. (2017). *Ecological Forecasting*. Princeton University Press.

4.  Gelman, A. & Dodhia, R. (2002). Let's practice what we preach: Turning tables into graphs.
