---
title: "Global Sensitivity Analysis Using Sobol Indices"
author: "Akash B V"
date: "`r Sys.Date()`"
format:
  html:
    self-contained: true
    df-print: paged
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    theme: cosmo
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

# Overview

This report presents results from a **variance-based global sensitivity analysis** of the SIPNET ecosystem model using Sobol indices. Unlike local sensitivity analysis (which measures rates of change at baseline conditions), variance-based methods decompose output variance into contributions from individual parameters and their interactions.

::: callout-note
## Why Sobol Indices?

Sobol indices provide a rigorous framework for:

-   **Quantifying parameter importance** via variance decomposition
-   **Detecting interactions** between parameters (non-additive effects)
-   **Factor prioritization** (which parameters to constrain for uncertainty reduction)
-   **Factor fixing** (which parameters can be fixed without loss of information)
:::

## Theoretical Background

**Sobol indices** are normalized variance contributions:

$$
S_i = \frac{V_i}{\text{VAR}(y)} \quad \text{(First-order index)}
$$

$$
T_i = 1 - \frac{\text{VAR}_{x_{\sim i}}[\mathbb{E}_{x_i}(y | x_{\sim i})]}{\text{VAR}(y)} = \frac{\mathbb{E}_{x_{\sim i}}[\text{VAR}_{x_i}(y | x_{\sim i})]}{\text{VAR}(y)} \quad \text{(Total-order index)}
$$

where $x_{\sim i}$ denotes all-parameters-but-$x_i$.

### Interpretation

-   $S_i$: Direct effect of $x_i$ on output variance (excluding interactions)
-   $T_i$: Total effect of $x_i$ including all interactions with other parameters
-   $T_i - S_i$: Interaction strength (purely higher-order effects)
-   $\sum S_i \approx 1$: Model is **additive** (linear-like behavior)
-   $\sum S_i \ll 1$: Model is **non-additive** (interactions dominate)

# Study Design

```{r}
#| label: setup
#| include: false

library(here)

# Tell here() this file is in analysis/ subdirectory
here::i_am("analysis/global_sensitivity.qmd")

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(DT)
library(stringr)
library(here)
library(config)
library(knitr)
library(scales)
library(ggthemes)

# Load configuration
cfg <- config::get(file = here::here("000-config.yml"))

# Load datasets
sobol_indices <- readr::read_csv(
  here::here(cfg$paths$data_dir, "sobol_indices.csv"),
  show_col_types = FALSE
)

metadata <- readRDS(here::here(cfg$paths$data_dir, "sobol_design_metadata.rds"))

# Extract metadata
N <- metadata$N
k <- metadata$k
total_runs <- metadata$total_runs
n_sites <- dplyr::n_distinct(sobol_indices$runid)
response_vars <- unique(sobol_indices$variable)

# Theme
theme_set(theme_bw(base_size = 12))
```

**Method:** Saltelli's sampling scheme with Jansen estimators

**Sample Size:** $N = `r N`$, Total Runs = `r total_runs`

**Scope:** Analysis performed across `r n_sites` sites

# Data Processing

We process the raw indices to parse namespaced parameters (e.g. grass.SLA), identify the Dummy parameter, and calculate interaction components.

```{r}
#| label: data-processing

# 1. Reshape and Clean
sobol_long <- sobol_indices %>%
  select(runid, variable, parameters, 
         Si_original, Si_low.ci, Si_high.ci,
         Ti_original, Ti_low.ci, Ti_high.ci) %>%
  mutate(
    # Enforce non-negativity (numerical artifacts can cause small negatives)
    Si = pmax(Si_original, 0),
    Ti = pmax(Ti_original, 0),
    # Interaction is the difference between Total and Main effects
    Interaction = pmax(Ti - Si, 0)
  )

# 2. Parse Parameter Names
sobol_parsed <- sobol_long %>%
  # Split "pft.param" into Prefix and Trait.
  # fill = "right" ensures "dummy" goes to Prefix, and Trait becomes NA.
  separate(parameters, into = c("Prefix", "Trait"), sep = "\\.", 
           extra = "merge", fill = "right") %>%
  mutate(
    # Handle Drivers/Dummy (where Trait is NA)
    Trait = ifelse(is.na(Trait), Prefix, Trait),
    
    # High-level Category for Partitioning
    Category = case_when(
      Prefix == "ic_ensemble" ~ "Initial Conditions",
      Prefix == "met_ensemble" ~ "Meteorology",
      Prefix == "dummy" ~ "Control (Noise)",
      TRUE ~ "Biological Parameter"
    ),
    
    # Nice Labels for plots
    Label = case_when(
      Prefix == "ic_ensemble" ~ "Init. Cond.",
      Prefix == "met_ensemble" ~ "Met. Driver",
      Prefix == "dummy" ~ "Dummy (Noise)",
      TRUE ~ paste(str_to_title(Prefix), Trait)
    )
  )

# 3. Aggregate Statistics across Sites
sobol_agg <- sobol_parsed %>%
  group_by(variable, Category, Label) %>%
  summarise(
    mean_Si = mean(Si, na.rm = TRUE),
    mean_Ti = mean(Ti, na.rm = TRUE),
    mean_Int = mean(Interaction, na.rm = TRUE),
    sd_Ti = sd(Ti, na.rm = TRUE),
    # Aggregate Confidence Intervals
    mean_Ti_low = mean(Ti_low.ci, na.rm = TRUE),
    mean_Ti_high = mean(Ti_high.ci, na.rm = TRUE),
    .groups = "drop"
  )

# Create summary (used in later sections)
sobol_summary <- sobol_agg
```

# Variance Decomposition (Source Partitioning)

Before ranking individual parameters, we assess the sources of uncertainty. We sum the First-Order indices ($S_i$) by category.

```{r}
#| label: fig-variance-partition
#| fig-cap: "Proportion of variance explained by Biology(param) vs. Environment(ic/met). The 'Control (Noise)' segment represents numerical approximation error."

variance_partition <- sobol_parsed %>%
  # Exclude the dummy from the partition plot to avoid confusion
  filter(Category != "Control (Noise)") %>%
  group_by(variable, Category) %>%
  summarise(Total_Si = sum(Si, na.rm = TRUE), .groups = "drop") %>%
  group_by(variable) %>%
  mutate(Percentage = Total_Si / sum(Total_Si))

ggplot(variance_partition, aes(x = variable, y = Total_Si, fill = Category)) +
  geom_col(position = "fill", width = 0.7) +
  geom_text(aes(label = percent(Percentage, accuracy = 1)), 
            position = position_fill(vjust = 0.5), 
            color = "white", fontface = "bold", size = 3) +
  scale_y_continuous(labels = percent) +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = "Variance Decomposition",
    y = "Proportion of Variance",
    x = "",
    fill = "Source"
  ) +
  theme_minimal()
```

# Parameter Ranking (First & Total Order)

This plot compares First-Order ($S_i$, solid bar) and Total-Order ($T_i$, outline) indices.

**Solid Bar (**$S_i$): Direct effect

**Gap (**$T_i - S_i$): Interaction effect

```{r}
#| label: fig-ranking-combined
#| fig-cap: "Top 15 inputs ranked by Total Effect (Ti). The solid colored bar represents the Main Effect (Si). The whiskers represent the Total Effect (Ti)."
#| fig-height: 10
#| fig-width: 10

# Filter top 15 by Ti (excluding dummy)
top_params <- sobol_summary %>%
  filter(Category != "Control (Noise)") %>%
  group_by(variable) %>%
  slice_max(order_by = mean_Ti, n = 15)

ggplot(top_params, aes(x = reorder(Label, mean_Ti), y = mean_Ti)) +
  # Total Order (Outline/Error Bar)
  geom_errorbar(aes(ymin = mean_Si, ymax = mean_Ti), width = 0.2, color = "grey50") +
  geom_point(aes(y = mean_Ti, color = Category), shape = 95, size = 10) +
  
  # First Order (Solid Bar)
  geom_col(aes(y = mean_Si, fill = Category), alpha = 0.8) +
  
  coord_flip() +
  facet_wrap(~variable, scales = "free_x") +
  scale_fill_tableau() +
  scale_color_tableau() +
  labs(
    title = "Parameter Importance: Main Effects vs. Total Effects",
    subtitle = "Solid Bar = Main Effect (Si); Top Whiskers = Total Effect (Ti)",
    x = "",
    y = "Sobol Index Value"
  ) +
  theme(legend.position = "bottom")
```

# Interaction Analysis

Parameters with large differences between $T_i$ and $S_i$ interact strongly with other factors.

```{r}
#| label: tbl-interactions

high_interaction <- sobol_summary %>%
  filter(Category != "Control (Noise)") %>%
  mutate(Interaction_Pct = (mean_Int / mean_Ti) * 100) %>%
  filter(mean_Int > 0.05) %>%
  arrange(desc(mean_Int)) %>%
  select(Variable = variable, Input = Label, 
         `Total (Ti)` = mean_Ti, `Main (Si)` = mean_Si, 
         `Interaction` = mean_Int, `Int. %` = Interaction_Pct)

DT::datatable(high_interaction, 
              caption = "Inputs with high interaction strength (> 0.05)",
              rownames = FALSE) %>%
  formatRound(columns = c(3:5), digits = 3) %>%
  formatRound(columns = 6, digits = 1)
```

# Factor Fixing (Screening)

To identify non-influential parameters, we compare their Total-Order index ($T_i$) against that of the Dummy Parameter (Noise). The dummy parameter varies randomly but has no effect on the model output.

Any parameter with $T_i$ indistinguishable from the Dummy (i.e., falling within the Dummy's confidence interval) can be safely fixed to a constant value.

```{r}
#| label: tbl-factor-fixing

# 1. Extract the Dummy Index (The Noise Floor) per variable
dummy_thresholds <- sobol_agg %>%
  filter(Category == "Control (Noise)") %>%
  select(variable, Noise_Ti = mean_Ti, Noise_Upper_CI = mean_Ti_high)

# 2. Compare All Parameters to the Dummy
fixable <- sobol_agg %>%
  left_join(dummy_thresholds, by = "variable") %>%
  filter(Category != "Control (Noise)") %>%
  # Criterion: Parameter Ti is less than the upper bound of the Noise
  filter(mean_Ti <= Noise_Upper_CI) %>%
  select(Variable = variable, Input = Label, `Mean Ti` = mean_Ti, `Noise Threshold` = Noise_Upper_CI) %>%
  arrange(Variable, `Mean Ti`)

if(nrow(fixable) > 0) {
  DT::datatable(fixable, 
                caption = "Parameters indistinguishable from numerical noise (Candidates for fixing)",
                extensions = 'Buttons', options = list(dom = 'Bfrtip', buttons = c('csv'))) %>%
    formatRound(columns = c(3:4), digits = 4)
} else {
  cat("**No parameters met the screening criteria.** All inputs exceed the numerical noise floor.")
}
```

# Summary Statistics

**Model Additivity:** The sum of First-Order indices ($\sum S_i$) indicates the degree of linearity. Values near 1.0 imply the model is additive (few interactions).

```{r}
#| label: calc-additivity

# Calculate model additivity (Sum of Si) per run
# We must sum across ALL parameters (excluding dummy) for each run
additivity_check <- sobol_parsed %>%
  filter(Category != "Control (Noise)") %>%
  group_by(runid, variable) %>%
  summarise(Sum_Si = sum(Si, na.rm = TRUE), .groups = "drop") %>%
  group_by(variable) %>%
  summarise(
    Mean_Additivity = mean(Sum_Si), 
    SD_Additivity = sd(Sum_Si),
    .groups = "drop"
  )

knitr::kable(additivity_check, 
             col.names = c("Variable", "Model Additivity (Mean Sum Si)", "SD"),
             digits = 3)
```

# Findings

**Drivers:** The Variance Decomposition plot highlights whether uncertainty is driven by intrinsic biology or extrinsic forcing.

**Additivity:** The mean sum of $S_i$ is `r round(mean(additivity_check$Mean_Additivity), 2)`.

**Screening:** The Factor Fixing table uses the Dummy parameter to rigorously identify irrelevant inputs.

# References

\[1\] Puy, A., et al. (2022). sensobol: An R Package to Compute Variance-Based Sensitivity Indices. *Journal of Statistical Software*, 102(5).

\[2\] Saltelli, A., et al. (2010). Variance based sensitivity analysis of model output. *Computer Physics Communications*.
