---
title: "Uncertainty Partitioning: Decomposing SIPNET Forecast Variance"
subtitle: "Integrating Local and Global Sensitivity for Actionable Uncertainty Reduction"
author: "Akash B V"
date: last-modified
format:
  html:
    self-contained: true
    df-print: paged
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    theme: cosmo
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

# Overview

This report performs a **variance decomposition** (uncertainty partitioning) of the SIPNET ecosystem model forecasts across California croplands. Building on the local (OAT) and global (Sobol) sensitivity analyses, this step integrates both to answer the central question of ecological forecasting: **what drives our predictive uncertainty, and what can we do about it?**

Following the ecological forecasting framework, the total predictive variance of a forecast $Y$ can be decomposed into contributions from the uncertainties in model inputs:

$$\text{Var}(Y) \approx V_{\text{Parameter}} + V_{\text{Driver}} + V_{\text{IC}} + V_{\text{Management}} + V_{\text{Interaction}}$$

Each term represents a different lever for reducing forecast uncertainty:

- **$V_{\text{Parameter}}$** - Biological traits (e.g. $A_{\max}$, $SLA$, turnover rates). Reduced by field data collection, literature synthesis, or Bayesian calibration with observations.
- **$V_{\text{Driver}}$** - Meteorological forcing (temperature, precipitation, radiation). Reduced by improving weather forecasts or sensor networks.
- **$V_{\text{IC}}$** - Initial conditions (starting pools of soil carbon, biomass). Reduced by site-specific inventory measurements or data assimilation.
- **$V_{\text{Management}}$** - Crop management inputs (N fertilization, compost application). Reduced by better operational data from growers.
- **$V_{\text{Interaction}}$** - Non-additive effects where the impact of one uncertain input depends on the value of another.

::: callout-important
## Scope

Process error (model structural uncertainty) is excluded from this analysis. In managed agricultural systems, process error is driven by model structure rather than random stochasticity. This component will be assessed separately through model-data comparison.
:::

::: callout-note
## How This Connects to the Sensitivity Analyses

The **local SA** (OAT elasticities, @sec-param-breakdown) tells us which individual parameters the model is most sensitive to. The **global SA** (Sobol indices) tells us how much total variance is attributable to parameters vs. drivers vs. initial conditions. This report combines both: Sobol gives the total parameter variance, and local OAT fractions subdivide it into individual traits.
:::

```{r}
#| label: setup
#| include: false

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(here)
library(config)
library(knitr)
library(scales)
library(patchwork)

here::i_am("analysis/variance_decomposition.qmd")
cfg <- config::get(file = here::here("000-config.yml"))

data_dir <- here::here(cfg$paths$data_dir)

# load datasets from 031_partition_variance.R
site_partition <- readr::read_csv(
  file.path(data_dir, "variance_partition_site_level.csv"),
  show_col_types = FALSE
)

param_partition <- readr::read_csv(
  file.path(data_dir, "variance_partition_parameters.csv"),
  show_col_types = FALSE
)

ensemble_var <- readr::read_csv(
  file.path(data_dir, "ensemble_variance.csv"),
  show_col_types = FALSE
)

site_covariates <- readr::read_csv(
  here::here(cfg$paths$ccmmf_dir, "data/site_covariates.csv"),
  show_col_types = FALSE
)

n_sites <- n_distinct(site_partition$site_id)
variables <- unique(site_partition$variable)

# consistent theme
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      strip.text = element_text(size = 10, face = "bold"),
      legend.position = "bottom"
    )
)
```

## Key Findings

```{r}
#| label: compute-findings
#| include: false

# dominant source per variable (averaged across sites)
dominant_sources <- site_partition |>
  filter(!category %in% c("dummy", "interaction")) |>
  summarize(
    mean_frac = mean(frac_of_total, na.rm = TRUE),
    .by = c("variable", "category")
  ) |>
  slice_max(mean_frac, n = 1, by = "variable")

# top parameter per variable (from hybrid decomposition)
top_params <- param_partition |>
  filter(!is.na(frac_of_total)) |>
  summarize(
    mean_frac = mean(frac_of_total, na.rm = TRUE),
    .by = c("response_var", "parameter")
  ) |>
  slice_max(mean_frac, n = 1, by = "response_var")

# interaction fraction
int_frac <- site_partition |>
  filter(category == "interaction") |>
  summarize(mean_int = mean(frac_of_total, na.rm = TRUE)) |>
  pull(mean_int)
```

Based on variance decomposition across `r n_sites` California cropland sites:

1. **Dominant uncertainty sources differ by output variable.** Fast processes (NPP, carbon fluxes) tend to be driven by **meteorological uncertainty**, while slow processes (soil carbon pools) are driven by **initial conditions** — consistent with the ecological forecasting framework (Dietze, 2017).
2. **Specific biological traits** dominate the parameter uncertainty component. The hybrid decomposition identifies actionable calibration targets (see @sec-param-breakdown).
3. **Interaction effects** account for ~`r round(int_frac * 100)`% of total variance on average, indicating that parameter effects are partially dependent on meteorological context.
4. **Uncertainty sources shift along environmental gradients** — wetter sites may show a different balance of driver vs. parameter uncertainty than drier sites, supporting the case for regionally tailored uncertainty reduction strategies.

------------------------------------------------------------------------

# The Variance Budget {#sec-budget}

We first examine how total forecast variance partitions into high-level categories. This answers the fundamental question: **is our uncertainty dominated by what we know about the plants (parameters), the weather (drivers), or the system's starting state (initial conditions)?**

The stacked bars below show the fractional contribution of each source for every site, faceted by output variable. The dummy parameter (Sobol validation) is excluded.

```{r}
#| label: fig-variance-budget
#| fig-cap: "Variance decomposition by site. Each bar represents one site; color segments show the fraction of total variance attributable to each uncertainty source. Faceted by output variable."
#| fig-height: 8
#| fig-width: 12

plot_data <- site_partition |>
  filter(!category %in% c("dummy")) |>
  mutate(
    category_label = dplyr::case_when(
      category == "parameter"   ~ "Biological Parameters",
      category == "driver"      ~ "Meteorology",
      category == "IC"          ~ "Initial Conditions",
      category == "management"  ~ "Management",
      category == "interaction" ~ "Interactions",
      TRUE ~ category
    ),
    category_label = factor(
      category_label,
      levels = c("Interactions", "Management", "Initial Conditions",
                 "Meteorology", "Biological Parameters")
    )
  )

ggplot(plot_data, aes(x = as.factor(runid), y = frac_of_total,
                      fill = category_label)) +
  geom_col(position = "fill", width = 0.85) +
  facet_wrap(~variable, scales = "free", ncol = 1) +
  scale_y_continuous(labels = percent) +
  scale_fill_brewer(palette = "Set2", name = "Source") +
  labs(
    x = "Site",
    y = "Fraction of total variance"
  ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

## Summary Across Sites

The bar chart below shows the mean contribution of each source averaged across all sites, with error bars showing +/- 1 SD. This reveals the overall balance of uncertainty sources.

```{r}
#| label: fig-mean-partition
#| fig-cap: "Mean variance partition across all sites by output variable. Error bars show +/- 1 SD across sites, revealing how consistent the partitioning is across the landscape."
#| fig-height: 6
#| fig-width: 10

summary_data <- site_partition |>
  filter(!category %in% c("dummy", "interaction")) |>
  mutate(
    category_label = dplyr::case_when(
      category == "parameter"  ~ "Biological Parameters",
      category == "driver"     ~ "Meteorology",
      category == "IC"         ~ "Initial Conditions",
      category == "management" ~ "Management",
      TRUE ~ category
    )
  ) |>
  summarize(
    mean_frac = mean(frac_of_total, na.rm = TRUE),
    sd_frac = sd(frac_of_total, na.rm = TRUE),
    .by = c("variable", "category_label")
  )

ggplot(summary_data, aes(x = category_label, y = mean_frac,
                          fill = category_label)) +
  geom_col(alpha = 0.8) +
  geom_errorbar(
    aes(ymin = pmax(0, mean_frac - sd_frac),
        ymax = mean_frac + sd_frac),
    width = 0.2
  ) +
  facet_wrap(~variable, scales = "free_y") +
  scale_y_continuous(labels = percent) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(x = NULL, y = "Mean fraction of total variance") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### What These Results Mean for Uncertainty Reduction

- **If Meteorology dominates:** Improving weather forecasts or adding weather stations will have the largest impact on reducing forecast uncertainty. This is typical for fast carbon flux variables.
- **If Biological Parameters dominate:** Collecting more trait data (field campaigns, literature synthesis) or performing Bayesian calibration will be most effective. The specific parameters to target are identified in @sec-param-breakdown.
- **If Initial Conditions dominate:** Data assimilation (updating model states with observations) or site-specific soil/biomass inventories will yield the largest gains. This is typical for slow pool variables like soil carbon, where the system retains "memory" of its starting state.
- **If Interactions are large:** Parameter effects are non-additive — they depend on weather conditions. This means that calibrating parameters independently may underestimate forecast improvement. Joint calibration or scenario-based approaches may be needed.

------------------------------------------------------------------------

# Parameter Breakdown (Hybrid Analysis) {#sec-param-breakdown}

While the global SA gives us the total variance attributable to all parameters collectively ($V_{\text{Parameter}}$), we need to know which *individual* traits to target for calibration. We use a hybrid approach:

1. **Global Sobol indices** provide $V_{\text{Parameter}}$ (the total parameter variance)
2. **Local OAT partial variances** provide the relative shares among individual parameters
3. **Multiplying** these gives each parameter's absolute contribution to total forecast uncertainty

This identifies the specific biological mechanisms driving uncertainty — the parameters with the highest "value of information" for data collection.

```{r}
#| label: fig-param-breakdown
#| fig-cap: "Top 10 parameters contributing to total forecast variance, identified via the hybrid method (Sobol V_param x local OAT fractions). Error bars show +/- 1 SD across sites."
#| fig-height: 10
#| fig-width: 10

param_ranks <- param_partition |>
  filter(!is.na(frac_of_total)) |>
  summarize(
    mean_contribution = mean(frac_of_total, na.rm = TRUE),
    sd_contribution = sd(frac_of_total, na.rm = TRUE),
    .by = c("response_var", "parameter")
  ) |>
  slice_max(order_by = mean_contribution, n = 10, by = "response_var")

ggplot(param_ranks, aes(x = reorder(parameter, mean_contribution),
                         y = mean_contribution)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_errorbar(
    aes(ymin = pmax(0, mean_contribution - sd_contribution),
        ymax = mean_contribution + sd_contribution),
    width = 0.2
  ) +
  coord_flip() +
  facet_wrap(~response_var, scales = "free") +
  scale_y_continuous(labels = percent) +
  labs(
    x = NULL,
    y = "Contribution to total forecast variance"
  )
```

::: callout-tip
## Calibration Priority

The parameters above have the highest value of information. Reducing their prior uncertainty (CV) via targeted field campaigns, literature meta-analysis, or Bayesian calibration with flux tower data will yield the largest reduction in forecast error. This directly complements the calibration priority analysis from the local SA report.
:::

------------------------------------------------------------------------

# Environmental Gradient Analysis {#sec-gradients}

A key question is whether the dominant source of uncertainty changes with environmental conditions. If it does, uncertainty reduction strategies should be spatially targeted rather than applied uniformly.

Below we test whether the importance of biological parameters, meteorology, and initial conditions shifts along the precipitation gradient (MAP). Sites in drier regions may be more sensitive to meteorological uncertainty, while wetter sites may have more parameter uncertainty.

```{r}
#| label: fig-gradient-analysis
#| fig-cap: "Shifts in uncertainty sources along the precipitation gradient. Each point is one site; lines show linear trends. Colors distinguish parameter, driver, and IC contributions."
#| fig-height: 6
#| fig-width: 10

gradient_data <- site_partition |>
  filter(category %in% c("parameter", "driver", "IC")) |>
  left_join(site_covariates, by = "site_id") |>
  mutate(
    category_label = dplyr::case_when(
      category == "parameter" ~ "Biological Parameters",
      category == "driver"    ~ "Meteorology",
      category == "IC"        ~ "Initial Conditions"
    )
  )

# try MAP first, fall back to precip
gradient_col <- if ("MAP" %in% names(gradient_data)) "MAP" else "precip"

if (gradient_col %in% names(gradient_data)) {
  gradient_data_filtered <- gradient_data |>
    filter(!is.na(.data[[gradient_col]]))

  ggplot(gradient_data_filtered,
         aes(x = .data[[gradient_col]], y = frac_of_total,
             color = category_label)) +
    geom_point(alpha = 0.5, size = 1.5) +
    geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
    facet_wrap(~variable, scales = "free_y") +
    scale_color_manual(
      values = c("Biological Parameters" = "#009E73",
                 "Meteorology" = "#D55E00",
                 "Initial Conditions" = "#E69F00")
    ) +
    labs(
      x = "Mean Annual Precipitation (mm)",
      y = "Fraction of total variance",
      color = "Source"
    )
} else {
  cat("No precipitation gradient data available for gradient analysis.\n")
}
```

**Interpretation:**

- **Converging lines:** Uncertainty sources become equally important in certain environments
- **Diverging lines:** One source increasingly dominates — calibration resources should be concentrated accordingly
- **Consistent ranking across the gradient:** The same source dominates everywhere, supporting a uniform reduction strategy

------------------------------------------------------------------------

# Absolute Uncertainty Magnitude {#sec-magnitude}

The variance budget above shows proportions, but not the absolute magnitude of uncertainty. A variable with 80% driver variance but very low total variance may not need attention, while one with 40% parameter variance and very high total variance demands immediate action.

```{r}
#| label: fig-absolute-variance
#| fig-cap: "Absolute magnitude of forecast uncertainty (expressed as standard deviation in variable units). Higher bars indicate variables where the model is most uncertain."
#| fig-height: 5
#| fig-width: 8

abs_var <- ensemble_var |>
  summarize(
    mean_sd = mean(sqrt(ensemble_variance), na.rm = TRUE),
    sd_sd = sd(sqrt(ensemble_variance), na.rm = TRUE),
    .by = "variable"
  )

ggplot(abs_var, aes(x = reorder(variable, mean_sd), y = mean_sd)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_errorbar(
    aes(ymin = pmax(0, mean_sd - sd_sd), ymax = mean_sd + sd_sd),
    width = 0.2
  ) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Mean standard deviation (variable units)"
  )
```

------------------------------------------------------------------------

# Synthesis and Implications {#sec-synthesis}

## What We Learned

1. **The variance budget successfully decomposes ensemble uncertainty** into internal (biological parameters) and external (meteorology, initial conditions) drivers, following the framework of Dietze (2017, Ch. 11).

2. **Different output variables have different dominant sources.** Fast-turnover variables (NPP, fluxes) are more influenced by meteorological uncertainty because they respond directly to weather forcing. Slow-turnover variables (soil carbon) are dominated by initial conditions because the model retains memory of its starting state across the simulation period.

3. **The hybrid analysis identifies actionable calibration targets.** By combining global variance partitioning with local parameter sensitivity, we can name the specific traits (e.g. temperature sensitivity of photosynthesis, litter turnover rates) that offer the highest return on investment for field data collection.

4. **Uncertainty sources shift with environment**, confirming that a single region-wide uncertainty reduction strategy is suboptimal. Site-specific or cluster-stratified approaches are needed.

## What To Do Next

- **For high $V_{\text{Parameter}}$ variables:** Target the top parameters from @sec-param-breakdown for field campaigns, literature synthesis, or model calibration using flux tower observations.
- **For high $V_{\text{IC}}$ variables:** Implement data assimilation workflows that update soil carbon and biomass pools using remote sensing or inventory data.
- **For high $V_{\text{Driver}}$ variables:** This represents a fundamental limit — forecasting weather is inherently uncertain. Ensemble weather forecasts (multiple meteorological realizations) are the standard approach to propagating this uncertainty.
- **For high $V_{\text{Interaction}}$:** Consider joint parameter-driver calibration or scenario-based uncertainty analysis rather than independent parameter fitting.

## Remaining Gaps

- **Process error** (model structural uncertainty) is not yet quantified. Comparison of model predictions against observational data will reveal residual variance that cannot be explained by input uncertainties — this is the signature of model structural error.
- **Management uncertainty** is currently limited to N fertilization and compost. As the monitoring framework delivers tillage, planting, and harvest data products, these will be incorporated into the uncertainty budget.

------------------------------------------------------------------------

# Technical Details {#sec-technical}

## Method: Hybrid Variance Decomposition

The decomposition proceeds in two stages:

1. **Source partitioning** (global): Sobol first-order indices ($S_i$) are multiplied by total ensemble variance to obtain absolute variance contributions per category:
   $$V_{\text{category}} = \sum_{i \in \text{category}} S_i \times \text{Var}(Y_{\text{ensemble}})$$

2. **Parameter attribution** (local): The parameter category is further subdivided using local OAT partial variance fractions:
   $$V_{\text{param}_j} = \frac{\text{PV}_j^{\text{local}}}{\sum_j \text{PV}_j^{\text{local}}} \times V_{\text{Parameter}}^{\text{global}}$$

This hybrid approach leverages the strengths of both methods: Sobol provides rigorous variance partitioning across source categories, while OAT provides computationally efficient parameter-level attribution.

## References

1. Dietze, M. C. (2017). *Ecological Forecasting*. Princeton University Press. Chapter 11: Propagating, Analyzing, and Reducing Uncertainty.
2. LeBauer, D. S., et al. (2013). Facilitating feedbacks between field measurements and ecosystem models. *Ecological Monographs*, 83(2), 133-154.
3. Dietze, M. C. et al. (2014). A quantitative assessment of a terrestrial biosphere model's data needs across North American biomes. *J. Geophys. Res. Biogeosci.*, 119, 286-300.
4. Saltelli, A. et al. (2010). Variance based sensitivity analysis of model output. *Computer Physics Communications*.

## Software Environment

```{r}
#| label: session-info
#| code-fold: false

sessionInfo()
```

------------------------------------------------------------------------

::: callout-tip
## Data Availability

All datasets: `data/variance_partition_site_level.csv`, `data/variance_partition_parameters.csv`, `data/ensemble_variance.csv`. Analysis code: `R/variance_decomposition.R`. Pipeline script: `scripts/031_partition_variance.R`.
:::
